---
title: "Getting started with CorOncoEndpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CorOncoEndpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CorOncoEndpoints)
```

## Describing a treatment group

A treatment group is described by `OncoArm()`. The inputs are the quantities
usually discussed at the design stage: the PFS median, the objective response
rate, the association between PFS and response, the proportion of PFS events
that are deaths, and the OS median.

```{r}
ctl <- OncoArm(pfs.median = 6, orr = 0.30, resp.cor = 0.40,
               death.prop = 0.15, os.median = 15, pps.hr.resp = 0.6,
               label = "Control")
ctl
```

The post-progression hazard of non-responders was calibrated to the OS median:
it is `r signif(ctl$gam0, 3)` per month, so the median post-progression survival
of non-responders is `r round(log(2) / ctl$gam0, 1)` months, and responders have
`r ctl$pps.hr.resp` times this hazard. The association with response was given
as a Pearson correlation; it can instead be given as the median PFS among
responders through `resp.pfs.median`.

The implied correlations among the three endpoints are

```{r}
CorEndpoints(ctl)
```

## A second group

The treatment group below has a PFS hazard ratio of 0.7 and a higher response
rate. Its post-progression hazard is set equal to that of the control group, so
the treatment acts on OS only through PFS and response.

```{r}
trt <- OncoArm(pfs.median = 6 / 0.7, orr = 0.45, resp.cor = 0.40,
               death.prop = 0.15, pps.hazard = ctl$gam0, pps.hr.resp = 0.6,
               label = "Treatment")
QuantileEndpoint(trt, endpoint = "os")
```

The implied OS median of the treatment group is
`r round(QuantileEndpoint(trt, endpoint = "os"), 1)` months.

## Generating data

`rOncoEndpoints()` generates patient-level data for many simulated trials. The
random numbers come from the `dqrng` generator; set `seed` for reproducible
results.

```{r}
dat <- rOncoEndpoints(nsim = 2, n = c(200, 200),
                      arms = list(Control = ctl, Treatment = trt),
                      a.time = c(0, 24), d.hazard = 0.005, seed = 1)
head(dat)
```

Every patient has `pfs_time <= os_time`, with equality when the PFS event is a
death (`progression = 0`). In this example
`r sum(dat$pfs_time > dat$os_time)` patients violate the ordering.

## Analysis at a cutoff

`EventTime()` gives the calendar time at which a number of events is reached in
each simulated trial, and `CutoffData()` returns the data observed at a cutoff.

```{r}
cut <- CutoffData(dat, EventTime(dat, events = 150, endpoint = "os"))
table(cut$sim, cut$os_event)
```

## Response observed at an interim analysis

With `resp.timing = "ttr"` each responder has a time to response, so the
response observed by an early cutoff can be derived. Responders also have PFS
longer than `resp.tau`, here the time of the first tumor assessment.

```{r}
arm_ttr <- OncoArm(pfs.median = 6, orr = 0.30, resp.cor = 0.40,
                   death.prop = 0.15, os.median = 15,
                   resp.timing = "ttr", resp.tau = 1.5, ttr.median = 2.5)
d_ttr <- rOncoEndpoints(nsim = 1, n = 300, arms = arm_ttr, a.time = c(0, 12),
                        seed = 2)
early <- CutoffData(d_ttr, cutoff = 9)
c(enrolled = nrow(early), responders_eventually = sum(early$response),
  responses_observed = sum(early$response_obs))
```
