---
title: "Model and theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model and theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CorOncoEndpoints)
```

## The model for one group

Let $(Z_1, Z_2)$ be standard bivariate normal with correlation $\theta$.

* PFS is $-\log\{\Phi(-Z_1)\}/\lambda_P$, so it is exactly exponential with
  hazard $\lambda_P$.
* Response is $R = 1$ when PFS exceeds the landmark $\tau$ (0 by default) and
  $Z_2 > c$, where $c$ gives $P(R = 1) = p$ exactly. The copula correlation
  $\theta$ is calibrated to the Pearson correlation between PFS and $R$ or to
  the median PFS of responders.
* A PFS event is a death with probability $\pi$ and otherwise a progression.
  After progression the patient survives for a time $D$, and
  $\mathrm{OS} = \mathrm{PFS} + \Delta D$, where $\Delta$ indicates progression
  as the first event. Hence $\mathrm{PFS} \le \mathrm{OS}$ for every patient.

Two models are available for $D$. In the illness-death model (`os.model =
"idm"`), $D$ is exponential with hazard $\gamma_0$ for non-responders and
$\gamma_1 = \kappa \gamma_0$ for responders (`pps.hr.resp` $= \kappa$). The
pre-progression death hazard is $\pi \lambda_P$, so the post-progression hazard
can be chosen separately from it. In the exponential-exponential model
(`os.model = "expexp"`), OS is exactly exponential.

## Why OS is not exponential in the illness-death model

When the pre-progression hazards are constant and the post-progression death
hazard stays bounded right after progression, both PFS and OS can be
exponential only if the post-progression hazard equals the pre-progression
death hazard. The proportion of PFS events that are deaths is then fixed at
$\pi = \mathrm{mPFS}/\mathrm{mOS}$, and the model is the maximal independence
model of Fleischer et al. (2009). For a PFS median of 6 months and an OS median
of 15 months this proportion is `r 6 / 15`.

`os.model = "expexp"` keeps both distributions exponential with a smaller
$\pi$ by letting the pre-progression death hazard decrease over time,
$\lambda_O e^{-c t}$. The post-progression hazard (in time since
randomization) is then determined by the exponential OS distribution:
$$
h_{12}(t) = \frac{\lambda_O e^{-\lambda_O t} - h_{02}(t) e^{-\lambda_P t}}
{e^{-\lambda_O t} - e^{-\lambda_P t}},
$$
so it is not a free input and cannot depend on response.

```{r}
ee <- OncoArm(pfs.median = 6, orr = 0.3, resp.cor = 0.4, death.prop = 0.15,
              os.model = "expexp", os.median = 15)
SurvEndpoint(ee, c(6, 12, 24), endpoint = "os")
exp(-log(2) / 15 * c(6, 12, 24))
```

## Decomposition of the correlations

In the illness-death model, $\Delta$ and $D$ are independent of PFS given
response. With $m_r = (1 - \pi)/\gamma_r$ and $\delta = m_1 - m_0$,
$$
\mathrm{Cov}(\mathrm{OS}, R) = \mathrm{Cov}(\mathrm{PFS}, R) + p(1-p)\delta, \quad
\mathrm{Cov}(\mathrm{PFS}, \mathrm{OS}) = \mathrm{Var}(\mathrm{PFS}) + \delta\,\mathrm{Cov}(\mathrm{PFS}, R).
$$
The partial covariance of OS and $R$ given PFS is
$\delta p (1-p) \{1 - \mathrm{Corr}(\mathrm{PFS}, R)^2\}$, so its sign is the
sign of $\delta$. Without a response effect on post-progression survival
($\kappa = 1$), $\mathrm{Corr}(\mathrm{OS}, R) = \mathrm{Corr}(\mathrm{PFS}, R)
\,\mathrm{Corr}(\mathrm{PFS}, \mathrm{OS})$.

```{r}
k1 <- OncoArm(pfs.median = 6, orr = 0.3, resp.cor = 0.4, death.prop = 0.15,
              os.median = 15)
k06 <- OncoArm(pfs.median = 6, orr = 0.3, resp.cor = 0.4, death.prop = 0.15,
               os.median = 15, pps.hr.resp = 0.6)
rbind(kappa_1 = CorEndpoints(k1), kappa_0.6 = CorEndpoints(k06))
```

With $\kappa = 1$, $\mathrm{Corr}(\mathrm{OS}, R)$ is
`r round(CorEndpoints(k1)[["cor.os.resp"]], 3)`, the product of
`r round(CorEndpoints(k1)[["cor.pfs.resp"]], 3)` and
`r round(CorEndpoints(k1)[["cor.pfs.os"]], 3)`. When responders also live
longer after progression ($\kappa = 0.6$), it rises to
`r round(CorEndpoints(k06)[["cor.os.resp"]], 3)`. The ordering
$\mathrm{Corr}(\mathrm{PFS}, R) \ge \mathrm{Corr}(\mathrm{OS}, R)$ is therefore
a consequence of the inputs and not an assumption.

## Attainable correlation between PFS and response

The Gaussian copula approaches the Frechet bounds as the copula correlation tends to -1 or 1.

```{r}
CorBoundPFSResponse(orr = 0.3)
CorBoundPFSResponse(orr = 0.3, pfs.median = 6, resp.tau = 1.5)
```

## References

Fleischer, F., Gaschler-Markefski, B. and Bluhmki, E. (2009). A statistical
model for the dependence between progression-free survival and overall
survival. *Statistics in Medicine*, 28, 2669--2686.

Meller, M., Beyersmann, J. and Rufibach, K. (2019). Joint modeling of
progression-free and overall survival and computation of correlation measures.
*Statistics in Medicine*, 38, 4270--4289.
